\documentclass[10pt]{article}

%\usepackage{html}


\title{ UAKGQuery: List of changes \newline
        \small{DocumentId:GradSof-UAKGQuery-eng-ch-17.04.2001-1C}
      }

\begin{document}

\maketitle{}

\tableofcontents

\section{  Changes in interface specifications. }

\begin{itemize}
 \item 1.0.2 - 1.1.0 
    \begin{itemize}
       \item added support for wstring.
       \item added support for binary large objects (BLOB-s)
       \item added method \verb|QueryManger::get_next_sequence_value|
       \item added method \verb|QueryManger::get_property|
    \end{itemize}
 \item 1.0.1 - 1.0.2 
    \begin{itemize}
      \item added logging command-line options.
      \item Changed order of initialization: added client library function
            \verb|initQueryService|
     \end{itemize}
 \item 1.0.0 - 1.0.1 IDL prefix tag is changed.
\end{itemize}

\section{  Changes of functional characteristics }

\begin{itemize}
 \item 1.0.2 - 1.1.0 none.
 \item 1.0.1 - 1.0.2 optional logging of SQL statements is added.
 \item 1.0.0 - 1.0.1 are absent.
\end{itemize}

\section{  Changes of implementation characteristics }

\begin{itemize}
 \item 1.0.2 - 1.1.0 
 \begin{itemize}
    \item Added support for InterBase back-end.
    \item Use version 1.4.0 of GradSoft C++ ToolBox
    \item Fixed known bugs of 1.0.2 version.
    \item Support of Orbacus-3.x and OmniORB-3 is stopped.
 \end{itemize}
 \item 1.0.0 - 1.0.2
 \begin{itemize}
    \item Use version 1.2.0 of GradSoft C++ ToolBox.
    \item Error reporting now use Logger.
    \item Eliminated initialization exception output, 
      when starting UAKQuweryService  with option \verb|--help|
 \end{itemize}
 \item 1.0.0 - 1.0.1 
 \begin{itemize}
    \item Fixed bug with incorrect marshalling of RCStream on machines
        with big-emdian byte order.
    \item non-transaction SQL sentences are commited immediatly after execution
        with OTS-enabled mode.
    \item review of ProgrammingGuide.
    \item fixed bug in process of compilation, when GradSoft C++ ToolBox headers are situated outside of default C++ include directory.
    \item fixed bug in description of installation process under Windows NT in Administration Guide.
 \end{itemize}
\end{itemize}

\section{  Changes in documentation }

\begin{itemize}
 \item 1.0.2 - 1.1.0 \begin{enumerate}
                     \item Administration guide:
                           \begin{enumerate}
                           \item description of installing procedure for Windows NT
                                 made to be adequate
                           \item pointed that transaction features for Windows NT are absent
                           \end{enumerate}
                     \item Programming guide:
                           \begin{enumerate}
                           \item routine of operating with LOB described
                           \item examples code review, InterBase to CORBA type relations described
                           \end{enumerate}
                     \end{enumerate}
 \item 1.0.0 - 1.0.2 - review
 \item 1.0.0 - 1.0.1 
 \begin{itemize}
   \item Few grammar bugs fixed.
   \item "List of changes" is added to documentation set.
   \item Formal document tracking attributes is added to documentation.
 \end{itemize} 
\end{itemize}

\section{  Changes in  distribution set }
 
\begin{itemize}
  \item 1.0.2 - 1.1.0 are absent
  \item 1.0.1 - 1.0.2 are absent
  \item 1.0.0 - 1.0.1 
 \begin{itemize}
   \item Documentation Set is changed.
 \end{itemize} 
\end{itemize}


\section{ List of deployed or published versions }

 \begin{itemize}
  \item 1.1.0
  \begin{itemize}
   \item begin of supplement - 28.01.2002
  \end{itemize}
  \item 1.0.2
  \begin{itemize}
   \item begin of supplement - 05.07.2001
   \item support discontinue - 01.03.2002 
   \item actions, must performed for depended software support: 
   \begin{itemize}
     \item recompile client applications with new libraries.
   \end{itemize}
  \end{itemize}
  \item 1.0.1
  \begin{itemize}
   \item begin of supplement - 07.05.2001
   \item support discontinue - 05.08.2001 
   \item actions, must performed for depended software support: 
   \begin{itemize}
     \item Upgrade you GradSoft C++ ToolBox to verion 1.2.0
     \item In client applications:
     \begin{itemize}
       \item insert call of initUAKGQueryService after ORB initialization.
       \item now if you do not use static global variable myORB, delete it.
     \end{itemize}
     \item recompile client applications with new libraries.
   \end{itemize}
  \end{itemize}
  \item 1.0.0
  \begin{itemize}
   \item begin of supplement - 17.04.2001
   \item support discontinue - 10.05.2001 
   \item actions on dependent software support: are absent. All interfaces specifications are fully compatible.
  \end{itemize}
 \end{itemize}

\end{document}


