

\documentclass[12pt]{article}
\usepackage[OT1,T2A]{fontenc}
\usepackage[koi8-u]{inputenc}
\usepackage{graphicx}
\usepackage{verbatim}

\title{ 
  UAKGQueryServicA: General Description
  \newline
  {\small Document Set: GOST 19.502-78 }
  \newline
  {\small DocumentID: GradSoft-UAKGQuery-e-G502-78-1999.4 }
}

\date{29.01.2002}
\author{ GradSoft }

\begin{document}

\maketitle{}

\tableofcontents

\section{  Introduction }

  UAKGQuery is a CORBA service for database access.

\section{ Intention:}

 \begin{enumerate}
   \item To provide universal CORBA API for database access.
   \item To provide integration of DB transactions into CORBA model of
 distributed transactions. 
 \end{enumerate}

\section{ Usage Description  }

  Application programmers use UAKGQueyry CORBA API for database access
 by linking application with client-side UAKGQuery libraries.
 Also embedding UAKGQuery Service into applications is supported, by linking
 it's with server-side libraries.

\section{Programming Environment Requirements}

 \begin{itemize}
  \item Database server - Oracle-8i,9i or InterBase-6.0 on any platform.
  \item Server-side operation system: Sun Solaris, Linux, Windows NT. 
  (port to any posix compabilite system is possible)
  \item Client-side operation system: Sun Solaris, Linux, Windows NT, FreeBSD 
  (port to any posix compabilite system is possible)
  \item ORB: ORBacus-4.x, TAO-1.1, TAO-1.2
  \item Transactions is supported when used ORBacus 4.x and OB OTS.
 \end{itemize}

\subsection{The requirments to the technical environment }

 \begin{itemize}
   \item The server part: the coincided with minimum queries target DB 
\item The client part: coincided with minimum queries target ORB 
\end{itemize}



\section{ Product Set }

\subsection{ Distribution Set }

   \begin{itemize}
     \item Source code.
     \item Source code of demos.
     \item Documentation Set.
   \end{itemize}

\subsection{ Executable and Libraries Set }

   \begin{itemize}
    \item Service as executable file
    \item Library for embedding server part of service into end-user application with collocated UAKGQuery calls
    \item Client side library
    \item Set of IDL filed, described access interfaces.
    \item Set of Facade C++ header files, for easy access to UAKGQuery data structures.
   \end{itemize} 
  
\subsection{ Documentation Set }

 \begin{enumerate}
   \item UAKGQuery: General Description. 
   \item UAKGQuery: Administation Guide.
   \item UAKGQuery: Programmer Guide.
   \item UAKGQuery: API description.
 \end{enumerate}

 General Description, Administration Guide and Programmer Guide are supplied
in pdf and html format in english and russian.

 API Guide is supplied in html format in english.

\section{ History of changes }

 \begin{itemize}
   \item 29.01.2002 - second edition.
   \item 06.06.2001 - review.
   \item 03.11.2000 - first edition.
   \item 14.12.2000 - review, translated into english.
   \item 22.12.2000 - file name changed, added DistributionSet.
 \end{itemize}

\end{document}
